A multi-scale story of the diffusion of a new technology: the web

### 2024-08-14 paper/submission1/tranos2023.qmd

---
title: "A multi-scale story of the diffusion of a new technology: the web"
author:
  - name: Emmanouil Tranos
    email: e.tranos@bristol.ac.uk
    affiliations: 
      - id: some-tech
        name: University of Bristol and The Alan Turing Institute
        department: 
        address: 
        city: 
        state: State
        postal-code: UK
    corresponding: e.tranos@bristol.ac.uk
address:
  - attributes:
        corresponding: true
    note: This is the first author footnote.
abstract: |
  This paper maps the participation in the digital economy and its evolution in the UK over space and time. Most of the existing economic geography literature which dealt with the spatiality of the internet employed supply-side measures, such as infrastructural capacity, in order to understand the geography of the digital economy and its potential spatial economic effects. Useful as these approaches might have been, they cannot capture the micro-processes and the characteristics of the individual online behaviour. Using large volumes of archived and geolocated web content, this paper models the diffusion of web technologies over space and time in the UK. The data and geolocation strategy allow to capture these processes at a very granular spatial scale. The modelling approach, which is based on simple spatial analytical methods and on the estimation of diffusion curves at various scales, enables to depict the role of geography and other cognitive factors which drove the diffusion of web technologies. Although the focus is on a recent historical period -- 1996-2012 -- the results of the analysis depict diffusion mechanisms which can be very useful in understanding the evolutionary patterns of the adoption of other newer technologies.
keywords: 
  - keyword1
  - keyword2
date: last-modified
bibliography: bibliography
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
editor: 
  markdown: 
    wrap: 72
---

@@

#insert libraries here
library(rprojroot)
library(tidyverse)
library(knitr)
library(kableExtra)

@@

# Introduction {#sec1}

@@

The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffused over space and time and use this knowledge to make predictions
regarding the diffusion of related *future* technologies. As per
@leibowicz2016representing, historical studies agree on the fact that
"technologies diffuse at different times, at different rates, and to
different extents in different places, and can be significantly
influenced by policies" [@victor1993]. @meade2021modelling highlight
that a variety of actors have a direct interest in gaining such
knowledge including network equipment suppliers; network operators,
regulatory and local authorities. These processes and their effects vary
a lot across scales: although the diffusion of a new technology might
not be optimal at a local level, it might be beneficial from a global
perspective as it could lead to faster diffusion to less advantaged
places [@leibowicz2016representing]. Despite the spatial heterogeneity
of such diffusion mechanisms and the policy relevance, there are very
limited attempts in the literature to analyse the diffusion of new
digital technologies at a detailed geographical level.

@@

Technological diffusion, which is by definition an aggregated process,
can be discussed in parallel with individual adoption mechanisms. On the
one hand, @rogers2010diffusion identifies early adopter of new
technologies as 'knowledgeable risk takers' and @griliches1957 as
'profit maximisers' [@ding2010modeling]. Such individual agents are
rewarded because of their attitude towards new technologies and
innovations. On the other hand, @perkins2011internet attribute diffusion
to two processes: (i) epidemic-like mechanisms, which are governed by
distance, proximity and social interactions, and (ii) by economic
mechanisms as new innovations are adopted by users as they become more
profitable, valueable and useful.
<!-- The criticisms are mainly based on the fact that although the approach -->
<!-- gives an idea of aggregate (industry or household) behaviour, it does -->
<!-- not focus on the individual's (firm or household) adoption process -->

@@

Importantly, the data used here depicts the *active* engagement with the
web in the UK as it contains all commercial websites that 
(i) are part of the UK's relevant second level domain (SLD, .co.uk), 
(ii) have been archived by the Internet Archive[^2], and 
(iii) include a mention to at least one valid UK
postcode in the web text. 
The act of creating a website is understood here as active engagement with the
web vis-à-vis the more passive act of browsing the web or having an
internet connection [@tranosuk]. Previous studies have focused mostly on
more passive notions of engaging with digital technologies such as
internet adoption and internet speeds [e.g. @blank2018local;
@riddlesden2014broadband]. More details about the data and the data
generation process can be found in Section [XX](#sec3).

@@

# Literature review {#sec2}

@@

Geographical diffusion is a synthesis of different processes. On the one
hand, we can identify purely spatial or, in other words, contagious
processes. Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovation just like pathogens spreads because of contagion and,
consequently, proximity and exposure [@hivner2003facilitating]. On the
other hand, we can identify hierarchical processes. Instead of
horizontal distance-based diffusion mechanisms, the top-down hierarchy
of urban systems drives diffusion. In reality, the synthesis of these
two processes represents how new technologies diffuse over space and
time [@morrill2020spatial].

@@

The temporal dimension was further explored by Everett Rogers and his
seminal work on 'Diffusion of Innovations' [@rogers2010diffusion].
Rogers being a sociologist, his work focused not on the diffusion of
innovations over space and time, but instead on the adoption of new
technologies and innovations by individuals and the individual
mechanisms that drive the decisions behind adoption. He identified five
groups of individuals regarding their adoption speed: innovators, early
adopters, early majority, late majority and laggards. The key mechanism of
diffusion and adoption is communication and how knowledge is
transferred within a social system. Therefore, all approaches agree on
the S-shaped diffusion and cumulative adoption pattern
[@grubler1990rise].

Schmidt's Law empirically illustrates a similar pattern. *Core* and
usually highly agglomerated regions is where new technologies are
invented and commercially deployed [@grubler1990rise].
This is where the first adopters tend to be based. Then, technologies
spread to the *rim* and eventually to the *periphery*. Although adoption
pace might be higher when new technologies finally arrive to the
periphery, the saturation levels there may never reach the ones in the
core because of the lack of infrastructure or other necessary
institutions [@leibowicz2016representing]. @grubler1990rise effectively
summarises the three key characteristics of the spatial diffusion
process: (i) the cumulative level of adoption follows an S-shaped
pattern just like purely temporal models; (ii) diffusion is shaped by a
hierarchy effect in a form of a centrifugal force: from core to
periphery; and (iii) diffusion is also shaped by distance and a
neighbourhood effect and contaminate nearby locations. These are the three
mechanisms that the empirical analysis in Section [4]{#sec4} investigates.

@@

# Materials and Methods {#sec3}

@@

To capture the diffusion of web technologies, a website density metric is developed 
for two different geographical scales: the Local Authority Districts (LAD) and the 
Output Areas (OA). The former is an administrative unit and there are 
\\textcolor{red}c. 380\\textcolor{black} such units in the UK. The latter is a census-based
geographical unit, which is very small as there are c. 230,000 of them in the UK.
This methodological choice will allow the mapping of the diffusion 
of web technologies and the assessment of the diffusion mechanisms 
at these two very different spatial scales.

The population of websites is calculated using data from the 
Internet Archive [^2] and, specifically, the JISC UK Web Domain Dataset [@ukwebarchive]. 
The Internet Archive is one of the most complete and oldest archive of webpages 
in the world operating since 1996 [@ainsworth2011much; @holzmann2016dawn].
It is a web crawler, which discovers webpages by following the hyperlinks
of every webpage its archives.
This dataset, which is curated by the British Library, contains all
the archived webpages from the UK ccTLD (.uk) from the 1996–2012 period. 
In essence, this is a long list of 2.5 billion URLs of archived webpages including 
also the archival timestamp.

@@

Instead of using the whole .uk country code top-level domain (ccTLD), this 
paper focuses on its commercial subset, the .co.uk SLD. 
This choice decrease the heterogeneity of the web data as 
such commercial websites have rather specific aims: they are used to diffuse information,
support online transactions and share opinions [@THELWALL2000441; @blazquez2018big]. 
Although a UK company can adopt a generic TLD such as.com and such cases escape 
the data used here, such omissions should not affect our results given the
popularity of the .uk ccTLD [@tranosuk]: UK consumers prefer to visit a .uk website
when they are searching for products or services [@hope]; and
anecdotal evidence indicates that during the first half of 2000, three
.co.uk domains were registered every minute [@oecd_coms]. 
Importantly, previous studies illustrated that .co.uk is the most popular UK SLD
[@tranosuk].

The text from these webpages was scanned using regular expressions (regex) to
identify strings of text which resemble UK postcodes and one fifth of them included
a mention to a postcode [@BL2013geo]. This information allows the geolocation of the 
data and the creation of the LAD and OA counts.

Before that, the data 
cleaning process included a data aggregation step, through which
the webpages were aggregated to the relevant websites. 
Based on the following example, all three webpages listed below are part of the 
same website ([http://www.website.co.uk](http://www.website.co.uk)) and at the end 
only websites and not the nested webpages were considered as otherwise the metric 
would have been biased towards large websites.

  - [http://www.website.co.uk/webpage_a](http://www.website.co.uk/webpage_a) B152TT

  - [http://www.website.co.uk/webpage_b](http://www.website.co.uk/webpage_b) BS81TH

  - [http://www.website.co.uk/webpage_c](http://www.website.co.uk/webpage_c) B152TT

\noindent What is challenging is that this aggregation approach,
which has been used elsewhere [@tranosuk; @shoreditch] may lead to websites with 
multiple postcodes. As per the above example, ([http://www.website.co.uk](http://www.website.co.uk))
includes two unique postcodes: B152TT and BS81SS. 
The distribution of postcodes for 2000 is presented in Table \ref{f2000}, which 
clearly illustrates the wide range. On one hand, we
have websites anchored to a unique location (72% of all
the reconstructed websites), which may represent
a small company with a single trading location. On the other hand,
we have websites with thousands of different postcodes.
Considering the timeperiod of the analysis, such cases can represent directories
which used to be popular in the pre-search engines early times of internet [@tranosuk].[^3]

The analysis presented here is based on two subsets of these data.
Firstly, the analysis is based on websites which only contain only one unique postcode
and, therefore, the geolocation process might suffer from less noise. 
As a robustness check, the analysis is replicated the analysis for an extended subset 
of websites, which include up to ten unique postcodes. These websites are geolocated 
by equally attaching them to multiple locations. This extended sample includes 
94 percent of all the archived websites in 2000.

@@

To create the web density metric, the yearly website counts at the LAD level are
standardised by the number of firms in LAD to avoid biases associated with LADs 
hosting a large number of firms. Given that there is no such statistic for the OA,
the actual OA level counts are used. However, because of the the rather
consistent spatial definition of OA (they host 40-250 households),[^4] 
website counts in OA are interpreted as a density metric too.

@@

This website metric...

@wilson201281 growth function description in p. 86

from R: Asym/(1+exp((xmid-input)/scal)) using the terms from Wilson
k/(1 + e\^(t_0 - t)/scal)

Wilson: k(1 + e\^-b(t - t_0)) So, scal = - 1/b

The curve is symmetric\* around t_0

**minimum R2 = 95%**, see also @grubler1990rise

The literature usually uses the saturation level as the asymptote. I am
using the total number of websites as we cannot compute a rate.

Moran's I as @ding2010modeling **TODO** for t_0, diffusion speed

# Results {#sec4}

1.  S-shaped diffusion curves: S for LADs per firm. UK,
    fast/slow/examples

@@

-   ideal: (i) train RF for all years and all (1) LADs and (2) OAs with
    CAST and report metrics. (ii) train for all years and all but one
    region for (1) LADs and (2) OAs to predict to the holdout region.
    Reports predictions as region similarities.

### RF results

The next section incorporates the above discussed spatial processes of the
diffusion of web technologies into the same modelling framework. The aim
is to use variables depicting these spatial processes in order to
predict the diffusion of web technologies in the UK over space and time
and across different scales. Specifically, four different models are
estimated. Firstly, all the data points at the OA and LAD are utilised
in order to build RF models and assess their capacity to predict the
adoption of web technologies. These two models will reveal the
predictive capacity of the spatial processes and also allow to see how
the importance of such variables changes across scales. The next two
sets of models will be again trained on web diffusion at the two working scales:
OA and LAD. However, instead of using all the data points, the OA and
the LAD from one of the twelve UK regions are held out and then the trained
model is used to predict web adoption in the OA or the LAD of the
held-out region. This process takes place recursively for all UK
regions. The difference in the predictive capacity of the different
samples will reveal how dissimilar are these spatial process across
regions and, importantly, at different scales.

>>>>

---
title: "A multi-scale story of the diffusion of a new technology: the web"
author:
  - name: Emmanouil Tranos
    email: e.tranos@bristol.ac.uk
    affiliations: 
      - id: some-tech
        name: University of Bristol and The Alan Turing Institute
        department: 
        address: 
        city: 
        state: State
        postal-code: UK
    corresponding: e.tranos@bristol.ac.uk
address:
  - attributes:
        corresponding: true
    note: This is the first author footnote.
abstract: |
  This paper maps the participation in the digital economy and its evolution in the UK over space and time. Most of the existing economic geography literature which dealt with the spatiality of the internet employed supply-side measures, such as infrastructural capacity, in order to understand the geography of the digital economy and its potential spatial economic effects. Useful as these approaches might have been, they cannot capture the micro-processes and the characteristics of the individual online behaviour. Using large volumes of archived and geolocated web content, this paper models the diffusion of web technologies over space and time in the UK. The data and geolocation strategy allow to capture these processes at a very granular spatial scale. The modelling approach, which is based on simple spatial analytical methods and on the estimation of diffusion curves at various scales, enables to depict the role of geography and other cognitive factors which drove the diffusion of web technologies. Although the focus is on a recent historical period -- 1996-2012 -- the results of the analysis depict diffusion mechanisms which can be very useful in understanding the evolutionary patterns of the adoption of other newer technologies.
keywords: 
  - keyword1
  - keyword2
date: last-modified
bibliography: bibliography
format:
  elsevier-pdf:
    keep-tex: true
    text: |
      \usepackage[demo]{graphicx} % "demo" option just for this example
      \usepackage{subcaption}
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
editor: 
  markdown: 
    wrap: 72
---

@@

#insert libraries here
library(rprojroot)
library(tidyverse)
library(knitr)
library(kableExtra)
library(sf)
library(cowplot)

@@

# Introduction {#sec-introduction}

@@

The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffused over space and time and use this knowledge to make predictions
regarding the diffusion of related *future* technologies. As per
@leibowicz2016representing, historical studies agree that technologies diffuse 
differently in terms of times, rates, and geographies and can be driven
by related policies [@victor1993]. @meade2021modelling highlight
that a variety of actors have a direct interest in gaining such
knowledge including network equipment suppliers; network operators,
regulatory and local authorities. These processes and their effects vary
a lot across scales: although the diffusion of a new technology might
not be optimal at a local level, it might be beneficial from a global
perspective as it could lead to faster diffusion to less advantaged
places [@leibowicz2016representing]. Despite the spatial heterogeneity
of such diffusion mechanisms and the policy relevance, there are very
limited attempts in the literature to analyse the diffusion of new
digital technologies at a detailed geographical level.

@@

Technological diffusion, which is by definition an aggregated process,
can be discussed in parallel with individual adoption mechanisms. On the
one hand, @rogers2010diffusion identifies early adopter of new
technologies as 'knowledgeable risk takers' and @griliches1957 as
'profit maximisers' [@ding2010modeling]. Such individual agents are
rewarded because of their attitude towards new technologies and
innovations. On the other hand, @perkins2011internet attribute diffusion
to two processes: (i) epidemic-like mechanisms, which are governed by
distance, proximity and social interactions, and (ii) economic
mechanisms as new innovations are adopted by users as they become more
profitable, valueable and useful.
<!-- The criticisms are mainly based on the fact that although the approach -->
<!-- gives an idea of aggregate (industry or household) behaviour, it does -->
<!-- not focus on the individual's (firm or household) adoption process -->

@@

Importantly, the data used here depicts the *active* engagement with the
web in the UK as it contains all commercial websites that 
(i) are part of the UK's relevant second level domain (SLD, .co.uk), 
(ii) have been archived by the Internet Archive[^2], and 
(iii) include a mention to at least one valid UK
postcode in the web text. 
The act of creating a website is understood here as active engagement with the
web vis-à-vis the more passive act of browsing the web or having an
internet connection [@tranosuk]. Previous studies have focused mostly on
more passive notions of engaging with digital technologies such as
internet adoption and internet speeds [e.g. @blank2018local;
@riddlesden2014broadband]. More details about the data and the data
generation process can be found in @sec-datamethods.

**S curve??**

@@

# Literature review {#sec-litreview}

@@

Geographical diffusion is a synthesis of different processes. On the one
hand, a purely spatial or, in other words, contagious processes can be identified. 
Adjacency and, more broadly, distance are the key drivers of
diffusion. This perspective draws similarities with epidemics:
innovation just like pathogens spreads because of contagion and,
consequently, proximity and exposure [@hivner2003facilitating]. On the
other hand, there is a hierarchical processes. Instead of
horizontal distance-based diffusion mechanisms, the top-down hierarchy
of urban systems shapes technological diffusion. In reality, the synthesis of these
two processes represents how new technologies diffuse over space and
time [@morrill2020spatial].

@@

The temporal dimension was further explored by Everett Rogers and his
seminal work on 'Diffusion of Innovations' [@rogers2010diffusion].
Rogers being a sociologist focused not on the diffusion of
innovations over space and time, but instead on the adoption of new
technologies and innovations by individuals and the individual
mechanisms that drive the decisions behind adoption. He identified five
groups of individuals regarding their adoption speed: innovators, early
adopters, early majority, late majority and laggards. The key mechanism of
diffusion and adoption is communication and how knowledge is
transferred within a social system. Therefore, all approaches agree on
the S-shaped diffusion and cumulative adoption pattern
[@grubler1990rise].

Schmidt's Law empirically illustrates a similar pattern. *Core* and
usually highly agglomerated regions is where new technologies are
invented and commercially deployed [@grubler1990rise].
This is where the first adopters tend to be based. Then, technologies
spread to the *rim* and eventually to the *periphery*. Although adoption
pace might be higher when new technologies finally arrive to the
periphery, the saturation levels there may never reach the ones in the
core because of the lack of infrastructure or other necessary
institutions [@leibowicz2016representing]. @grubler1990rise effectively
summarises the three key characteristics of the spatial diffusion
process: (i) the cumulative level of adoption follows an S-shaped
pattern just like purely temporal models; (ii) diffusion is shaped by a
hierarchy effect in a form of a centrifugal force: from core to
periphery; and (iii) diffusion is also shaped by distance and a
neighbourhood effect and contaminate nearby locations. These are the three
mechanisms that the empirical analysis in @sec-results investigates.

@@

# Data and Methods {#sec-datamethods}

@@

<!-- main metric -->

To capture the diffusion of web technologies, a website density metric is developed 
for two different geographical scales: the Local Authority Districts (LAD) and the 
Output Areas (OA). The former is an administrative unit and there are 
c. 374 such units in the UK. The latter is a census-based geographical unit, which 
is very small as there are c. 230,000 of them in the UK. This methodological choice 
will allow the mapping of the diffusion of web technologies and the assessment of 
the diffusion mechanisms at these two very different spatial scales.

<!-- web data -->

The counts of websites at these scales are calculated using data from the 
Internet Archive [^2] and, specifically, the JISC UK Web Domain Dataset [@ukwebarchive]. 
The Internet Archive is one of the most complete and oldest archive of webpages 
in the world operating since 1996 [@ainsworth2011much; @holzmann2016dawn].
It is a web crawler, which discovers webpages by following the hyperlinks
of every webpage its archives.
This dataset, which is curated by the British Library, contains all
the archived webpages from the UK ccTLD (.uk) from the 1996–2012 period. 
In essence, this is a long list of 2.5 billion URLs of archived webpages including 
also the archival timestamp.

@@

Instead of using the whole .uk country code top-level domain (ccTLD), this 
paper focuses on its commercial subset, the .co.uk SLD. 
This choice decrease the heterogeneity of the web data as 
such commercial websites have rather specific aims: they are used to diffuse information,
support online transactions and share opinions [@THELWALL2000441; @blazquez2018big]. 
Although a UK company can adopt a generic TLD such as .com and these cases escape 
the data used here, such omissions should not affect our results given the
popularity of the .uk ccTLD [@tranosuk]: UK consumers prefer to visit a .uk website
when they are searching for products or services [@hope]; and
anecdotal evidence indicates that during the first half of 2000, three
.co.uk domains were registered every minute [@oecd_coms]. 
Importantly, previous studies illustrated that .co.uk is the most popular UK SLD
[@tranosuk].

The text from these webpages was scanned using a regular expression (regex) to
identify strings of text which resemble UK postcodes and one fifth of them included
a mention to a postcode [@BL2013geo]. This information allows the geolocation of the 
data and the creation of the LAD and OA counts.

The data cleaning process included an aggregation step, through which
the archived webpages were aggregated to the parent websites. This website 
reconstruction allows the creation of a *website* instead of a *webpage* density 
metric. Websites, which are hierarchical structures, tend to represent specific 
orgnanisations or entities, and, arguably, are more meaningful observational units 
than webpages, which ignore the upstream dependencies. 
Based on the following example, all three webpages are part of the 
same website ([http://www.website.co.uk](http://www.website.co.uk)) and at the end 
only websites and not the nested webpages were considered as otherwise the metric 
would have been biased towards large, place specific, websites.

  - [http://www.website.co.uk/webpage_a](http://www.website.co.uk/webpage_a) B15 2TT

  - [http://www.website.co.uk/webpage_b](http://www.website.co.uk/webpage_b) BS8 1TH

  - [http://www.website.co.uk/webpage_c](http://www.website.co.uk/webpage_c) B15 2TT

\noindent What is challenging is that this aggregation approach,
which has been used elsewhere [@tranosuk; @shoreditch] may lead to websites with 
multiple postcodes. As per the above example, [www.website.co.uk](http://www.website.co.uk)
includes two unique postcodes: B15 2TT and BS8 1SS. 
The distribution of postcodes for 2000 is presented in Table \ref{f2000}, which 
clearly illustrates the wide range. At the left end of the distribution, there are 
websites anchored to a unique location (72% of all the reconstructed websites in
2000), which may represent a small company with a single trading location. At the 
right end, we have websites with thousands of different postcodes.
Considering the time period of the analysis, such cases can represent directories
which used to be popular in the pre-search engines early times of the commercial 
internet [@tranosuk].[^3]

The analysis presented here is based on two subsets of these data. Firstly, on websites, 
which only contain only one unique postcode and, therefore, the geolocation process 
does not suffer from noise. As a robustness check, the analysis is replicated 
for an extended subset of websites, which include up to ten unique postcodes. These 
websites are geolocated by equally attaching them to multiple locations. This extended 
sample includes 94 percent of all the archived websites in 2000.

@@

Another data cleaning step dealt with some extreme outliers. Figure \ref{pc1} plots
the website counts for the top 1000 postcodes. Some obvious outliers can be observed 
for the 2004-2006 period, which can be attributed to a link farm [@BL2013links].  
The website counts for these six postcodes (SE24 9HP, CV8 2ED, GL16 7YA, CW1 6GL, 
M28 2SL, DE21 7BF), which in 2004 or 2005 had more than 1000 websites pointing to 
them, were replaced with predicted values based on a simple panel regression model 
with postcode fixed effects and yearly dummy variables. As the @sec-results illustrates,
this led to a small increase of the model predictive capability. To put the magnitude 
of the data imputation into perspective, this process affected 6 out of the 557,808
postcodes present in the data.

```{r echo=FALSE, message=FALSE, fig.cap="\\label{pc1}Yearly website counts per postcode (top 1000)", out.height="60%"}
path.image <- paste0(path, "/outputs/pc_year_1.png")
knitr::include_graphics(path.image)
```

To create the website density metric, the yearly website counts at the LAD level are
standardised by the number of firms in LAD to avoid biases associated with LAD 
hosting a large number of firms. Given that there is no such statistic for the OA,
the actual OA level counts are used. However, because of the the consistent spatial 
definition of OA (they host 40-250 households),[^4] website counts in OA are 
interpreted as a density metric too.

@@

<!-- system-level analysis -->

These website density metrics are used for the three different steps of the analysis. 
Firstly, a system-level analysis explores how the diffusion of web technologies in
the UK fits with the well-established S-curve. To do so, the following logistic 
function (Equation \ref{eq:s}) is estimated for the whole of the UK and for each 
LAD separately:

\begin{align}
y = k /(1 + e^{-(t-t_{o})})\label{eq:s}
\end{align}

\noindent $k$ is the asymptote or, in other words, the saturation level, $b$ the overall 
growth rate, and $t_{0}$ the *inflection point*
of maximum growth at $k/2$, where the logistic function is symmetrical [@wilson201281]. 
We compare  the LAD $t_0$ of each LAD against the $t_0$ of the UK to delineate 
whether a LAD reached that point faster or slower than the country average. Importantly, 
an accuracy criterion was imposed and only S-curves with $R^2 > 0.9$ were included 
in the analysis. To estimate Equation \ref{eq:s} a self-starting logistic growth 
model was employed using the `nls` and `SSlogis` functions in `R`.

<!-- **minimum R2 = 95%** in @wilson201281, see also @grubler1990rise -->
<!-- The literature usually uses the saturation level as the asymptote. I am -->
<!-- using the total number of websites as we cannot compute a rate. -->

The system-level analysis also focuses on the volatility of the web adoption to 
depict places with high concentration of early adopters and, equally, latecomers. 
To do so, the change of the ranking of the UK LAD over time is plotted and discussed. 
Both the S-curve and the volatility analysis focus only on LAD as the very large 
number of OA would have made such analysis diffucult to visualise and interpret.

<!-- ESDA -->

Next, exploratory spatial data analysis depicts whether the two main 
drivers of spatial diffusion -- namely neighbourhood and hierarchy -- underpin the
diffusion of web technologies in the UK. To capture the former and following 
@ding2010modeling the Moran's I and the Local Index of Spatial Association (LISA) 
are estimated for website density. To address the hierarchy mechanism, the Gini 
coefficient -- a well established metric of inequality -- is calculated. All the 
above are computed and plotted longitudinally both for the LAD and the OA.

<!-- RF -->

Lastly, a modelling framework is developed to test the above diffusion mechanisms.
The overarching aim is to build a model that can test the relationship between the
website density and these mechanisms:

\begin{align}
Website\,Density_{t} \sim Hierarchy_{t-1} + Neighbourhood_{t-1} + S-curve_{t}\label{eq:rf.generic}
\end{align}

To estimate Equation \ref{eq:rf.generic}, Random Forest (RF) models were built. 
This is a popular machine learning algorith for both regression and classification 
problems [@biau2012analysis]. It was introduced by @breiman2001random and has  
gained popularity, becoming a go-to data science tool. RF can effectively 
handle skewed distributions and outliers, model non-linear relationships, require 
minimal hyperparameter tuning, exhibit low sensitivity to these parameters, and 
have relatively short training times [@Caruana2008; @liaw2002classification; @yan2020using]. 
These attributes match well with the website density data characteristics including 
skewness especially for OA. Also, the large data size (c. 230k data points for 
each of the 17 years) call for fast training times. 
Importantly, RF predictions tend to be more accurate than those from single regression 
trees and outperform Ordinary Least Squares in out-of-sample predictions, even with 
moderate-sized training data and a small number of predictors [@mullainathan2017machine; 
@athey2019machine; @sulaiman2011intelligent; @pourebrahim2019trip; @biau2012analysis].

RF is a tree-based ensemble learning algorithm [@breiman2001random]. It begins by 
generating random samples of the training data, which are then used to grow 
regression trees to predict the dependent variable. Data points and predictors are 
randomly sampled for the different trees. The trees are trained in parallel using their 
own bootstrapped samples of the training data. A crucial feature of RF is their 
ability not to overfit, meaning they can generalize well to unseen test data. While 
each tree may overfit individually, the ensemble of trees does not because the errors 
of individual trees are averaged, reducing the overall variance and preventing overfitting 
[@last2002improving]. For regression problems, RF predictions are made by averaging 
the predictions of all decision trees.

RF have been widely employed to address regression research problems. @pourebrahim2019trip 
combined a spatial interaction modeling framework with ML algorithms including RF
to predict commuting flows in New York City. @sinha2019assessing advocated for adopting
spatial ensemble learning approaches, such as RF, to model spatial data with high
autocorrelation and heterogeneity. @creditspatial predicted employment density in 
Los Angeles using spatially explicit RF. @guns2014recommending used RF to build a 
recommendation system for research collaborations. @ren2019predicting trained RF 
to predict the socio-economic status of cities using various online and mobility 
predictors. @tranos2023using utilised hyperlinks data and RF to make out-of-sample
predictions of interregional trade. @zhou2023geography employed such a framework
to assess whether key predictors of obesity differ across English cities.

# Results {#sec-results}

1.  S-shaped diffusion curves: S for LAD per firm. UK,
    fast/slow/examples

@@

-   ideal: (i) train RF for all years and all (1) LAD and (2) OA with
    CAST and report metrics. (ii) train for all years and all but one
    region for (1) LAD and (2) OA to predict to the holdout region.
    Reports predictions as region similarities.

<!-- S-shaped -->

Figure \ref{s_uk} plots the S curve for the cumulative adoption of website technologies
in the UK during the 1996-2012 period. It demonstrates a pattern well aligned with
previous studies discussed in @sec-litreview. The vertical line for year 
2003 illustrates the point where the modeled cumulative adoption was equal to 50% 
of the maximum. This $t_0$ *inflection* point signals the maximum adoption speed 
and is used here to determine whether the UK LAD reached that inflection point 
earlier or later than the UK average. Specifically, the S curve and the inflection 
point are estimated for every LAD individually and then compared with the UK average. 
LAD that reached their inflection point earlier than the country average are labelled 
as *fast* and the rest as *slow*. Figure \ref{s_map} maps this pattern and the picture 
is not entirely clear. On the one hand, there are some expected examples of LAD
with relevant industrial backgrounds delineated as fast: the City of London, a 
world-renowned cluster of finance industries [@cook2007role], and Reading, a town 
with high-tech service industries in proximity to London and its main airport, Heathrow 
[@england2005polynet]. On the other hand, LAD which were expected to appear as fast
-- e.g. Hackney in central London and Bristol, a well-established creative cluster 
[@oatley1999cultural; @bassett2002cultural] -- were delineated as slow. Nevertheless, 
out of the 10 fastest LAD, eight were located in Greater London and one in 
Cambridge (see **Appendix**), but in overall there is a spatially heterogenous and
not easy to explain spatial pattern.

```{r echo=FALSE, message=FALSE, fig.cap="\\label{s_uk}Grwoth curve, UK"}
path.image <- paste0(path, "/outputs/s/s_uk_per_firm.png")
knitr::include_graphics(path.image)
```

```{r echo=FALSE, message=FALSE, fig.cap="\\label{s_map}LAD adoption rates"}
path.image <- paste0(path, "/outputs/s/speed_map.png")
knitr::include_graphics(path.image)
```

<!-- rankings -->
The next step is to assess the stability and volatility of the LAD in terms of 
their adoption of web technologies. As we know from the literature [@risk_perceptions],
different agents have different perceptions about and levels of acceptance of the 
risks and the potential economic returns associated with the adoption of new technologies 
-- see for instance the seminal work of @venkatesh2000theoretical. To reveal such 
aggregated patterns, Figures \ref{rank_stable} and \ref{rank_unstable} plot the ranking
of UK LAD based on website density. To decrease noise, the average ranking 
of 1996-1998 and 2010-2012 is plotted instead of the individual years. 
While there are quite a few obvious cases of LAD that maintained their position 
between the beginning and the end of the study period either at the top or at the 
bottom of the hierarchy (Figure \ref{rank_stable}), there are also 
quite a few LAD that changed drastically their position. Some of these LAD enjoyed
a process that at the first instance looks like leapfrogging since 
they managed to jump at the top of the hierarchy despite their slow start (top of 
Figure \ref{unstable}). There is extensive literature regarding the potential benefits of
technological leapfrogging. The underpinning argument is that latecomers can adopt
and benefit from new technologies that have been developed elsewhere without incurring 
the hefty initial R&D costs [@teece2008firm]. Although the leapfrogging literature
does not pay much attention to cities and regions [@yu2018sustainability], previous 
research highlighted the long term and sustained productivity benefit of the early
adopters of web technologies in the UK [@tranosuk]. So, although the LAD system 
is volatile, it is not clear whether the LAD with high concentration of late adopters 
will gain any latecomer benefits in a way similar to countries experiencing such
technological leapfrogging.

```{r echo=FALSE, message=FALSE, fig.cap="\\label{rank_stable}Dynamics of wed diffusion: stability of LAD"}
path.image <- paste0(path, "/outputs/ranks/web_per_firm2000_2012_only05_av.png")
knitr::include_graphics(path.image)
```

```{r echo=FALSE, message=FALSE, fig.cap="\\label{rank_unstable}Dynamics of wed diffusion: volatility LAD"}
path.image <- paste0(path, "/outputs/ranks/web_per_firm2000_2012_only95_av.png")
knitr::include_graphics(path.image)
```

<!-- Neighbourhood effect -->

Then, figures \ref{morani} and \ref{lisa} offer a first insight into whether a neighbourhood 
effect underpins the diffusion of web technologies in the UK as they plot the Moran's I
and the LISA maps of website density respectively for LAD and OA. Starting from 
the former, spatial autocorrelation was higher in the beginning and then around 
year 2000 dropped slightly and stabled at 0.2. This reflects the early concentration 
of high website density around London, which over time diffused as high-high clusters
can be seen in other parts of the country away from London as seen in Figure \ref{lisa}. An 
almost reverse pattern can be observed for OA. At the beginning of the study period
Moran's I was around 0.1 and it plateaued after 2000-2001 around 0.2. Because of the 
very small size of OA, at the early stages of the diffusion of web technologies 
their adoption was spatially scattered. This is reflected in the lack of any significant
clusters in 1996. Eventually, as the adoption rate increased, more such clusters 
of high website density were formed and this is reflected both in the Moran's I 
and the LISA maps in Figures \ref{morani} and \ref{lisa}. All in all, the exploratory 
spatial data analysis advocates towards an underpinning neighbourhood effect and 
the different scales of analysis illustrate how it evolved differently over time.

```{r, morani, echo=FALSE, message=FALSE, fig.cap="\\label{morani}Website density Moran's I"}

path.in.la <- paste0(path, "/outputs/lisa/corrected/LA/morani_la.csv")
path.in.oa <- paste0(path, "/outputs/lisa/corrected/oa/morani_oa.csv")

la.m <- read_csv(path.in.la) %>% 
  filter(p < 0.001) %>% # all values were significant
  mutate(spatial.unit = 'LAD')
oa.m <- read_csv(path.in.oa) %>% 
  filter(p < 0.001) %>% # all values were significant
  mutate(spatial.unit = 'OA')

bind_rows(la.m, oa.m) %>% 
  ggplot(aes(x=year, y=morani, fill= spatial.unit)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  ylab('Moran\'s I') +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_manual('Spatial unit', 
                    values = c("LAD" = "grey80",
                                "OA"="grey48"),
                     limits = force) +
  scale_x_continuous(NULL, labels = as.character(1996:2012), breaks = 1996:2012) 
```

Figure \ref{lisa}

```{r, echo=FALSE, message=FALSE, fig.cap="\\label{lisa}\\centering Website density LISA maps", out.height="80%"}
